


# Chapter 6: Analysis of longitudinal data
Mounir Ould Setti
09/12/2019

## I - RATS data (Chapter 8 of MABS)

### 1. Loading the libraries, importing the RATSL dataset, and summary of the data
```{r eval=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)
```

```{r}
RATSL<- read.table("data/RATSL.csv", header = TRUE, sep = ',')
RATSL$X <- NULL
RATSL$ID <- as.factor(RATSL$ID)
RATSL$Group <- as.factor(RATSL$Group)
```



```{r}
glimpse(RATSL)
summary(RATSL)
str(RATSL)

```

We have taken the time on the wrangling phase of this exercise to turn the RATS dataset into a longitudinal form.

The long dataset form has more rows and less columns. A subject can have many rows represeting their observations at different time points which are indicated in the time column.

The study design associated to the dataset seems that it assigned the rats into three (intervention) groups and then observations of weight have been recorded on each rat at different time points.

### 2. Graphical representation of the data

``` {r}
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

```

Figure 1. Individual rats weight development by group for the RATS data

It seems that rats in the three groups gain weight with time but the groups 2 and 3 seem to gain weight more markedly. However, the difference between the groups might possibly be due to the differences at baseline.
Standardization might help fix this issue.

Standardizing the variable Weight

``` {r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = ((Weight-mean(Weight))/sd(Weight))) %>%
  ungroup()
#Glimpse the data
glimpse(RATSL)

```


Ploting again with the standardized weights

``` {r}
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +  theme(legend.position = "none") +
  scale_y_continuous(name = "standardized weight")
```

Figure 2. Individual rats weight development by group for the RATS data after standardization

We notice a decline in weight development over time in the three groups.

### 3. Summary measure analysis
It could be helpful in representing the data to summarize observations into their mean accompanied by standard error.

Summarizing the data

``` {r}
# Time, baseline included
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of RATSL by Group and Time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

```

Plotting the mean weights

``` {r}
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

Figure 3. Mean weights for the three groups in the RATS data

We notice that the three groups have distinctive weights development over time. Their confidence intervals do not overlap. 

Creating a summary data by group and rat with mean as the summary variable (ignoring baseline time 1) and plotting it:

```{r}
boxplot(mean~Time,data=RATSS,main="Rats weights data",xlab="Time",ylab="Mean weights")

```

Figure 4. Boxplots for the times ofthe RATS data.

Applying a summary measure approach might help.

From figure 3 and the boxplot on figure 4, it looks like our data is of type Growth (Table 8.2 of MABS) where the eventual value of outcome is different between groups. The recommended summary measure approach is to use the percentage of change between first and last values.

As first values I will use the first quartile, and as last values, I will use the 3rd quartile.

As percentage of change between the lasts and the firsts, I will use the mean of the first quartile and the mean of the last quartile.



``` {r}
# Create a summary data by treatment and subject with percentage of change between first and last values as the summary variable
RATSS2 <- RATSL %>%
  group_by(Group, ID) %>%
  summarise( diffperc= ((mean(Weight[Time>49]) /
                            mean(Weight[Time<16])
                            )-1)*100 ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS2)

# Draw a boxplot of the percentage of difference versus group
ggplot(RATSS2, aes(x = Group, y = diffperc)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "Percentage of difference, Time")

```

It seems that group 2 has the highest gain in weight with about 12% weight gain between the weight in the first quartile of time and the weight at the third quartile of time (percentage of difference between baseline and end-measures).
However, the confidence intervals overlap, suggesting that differences between groups might not be statistically significant.

A statistical test might help evaluating the differences between groups.

A one way analysis of variances might be of help to test for differences between three groups.

``` {r}
summary(aov(diffperc ~ Group, data = RATSS2))

```

It seems that, indeed, the groups do not differ significantly in terms of weight gain over time.


## II - BPRS dataset (Chapter 9 of MABS)