---
title: "Chapter 3: Logistic Regression"
author: "Mounir Ould Setti"
date: "18/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 2. Exploring the data

Importing the data from the URL
``` {r}
alc <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/alc.txt",sep=",",header = TRUE)

colnames(alc)
str(alc)
dim(alc)
```

The data explores students' performance along with some behavioral parameters such as alochol use. It consists of 382 observations corresponding to the students and 35 variables.

## 3. Hypotheses on the association of alcohol consumption with some other variables

1- Study time (variable 'studytime): I suppose that high alcohol use would be associated with low study times.

2- Class failures (variable 'failures'): I suppose that students who have low alcohol use would be likely not to have class failures.

3- Sex: I suppose that men would be more likely to have high use of alcohol

4- Absences: I suppose that those with high alcohol use have a higher count of absences.

## 4. Exploring the distribution of my chosen variables and their relationships with alcohol consumption

