


# Chapter 6: Analysis of longitudinal data
Mounir Ould Setti
09/12/2019

## I - RATS data (Chapter 8 of MABS)

### 1. Loading the libraries, importing the RATSL dataset, and summary of the data
```{r eval=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)
```

```{r}
RATSL<- read.table("data/RATSL.csv", header = TRUE, sep = ',')
RATSL$X <- NULL
RATSL$ID <- as.factor(RATSL$ID)
RATSL$Group <- as.factor(RATSL$Group)
```



```{r}
glimpse(RATSL)
summary(RATSL)
str(RATSL)

```

We have taken the time on the wrangling phase of this exercise to turn the RATS dataset into a longitudinal form.

The long dataset form has more rows and less columns. A subject can have many rows represeting their observations at different time points which are indicated in the time column.

The study design associated to the dataset seems that it assigned the rats into three (intervention) groups and then observations of weight have been recorded on each rat at different time points.

### 2. Graphical representation of the data

``` {r}
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

```

Figure 1. Individual rats weight development by group for the RATS data

It seems that rats in the three groups gain weight with time but the groups 2 and 3 seem to gain weight more markedly. However, the difference between the groups might possibly be due to the differences at baseline.
Standardization might help fix this issue.

Standardizing the variable Weight

``` {r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = ((Weight-mean(Weight))/sd(Weight))) %>%
  ungroup()
#Glimpse the data
glimpse(RATSL)

```


Ploting again with the standardized weights

``` {r}
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +  theme(legend.position = "none") +
  scale_y_continuous(name = "standardized weight")
```

Figure 2. Individual rats weight development by group for the RATS data after standardization

We notice a decline in weight development over time in the three groups.

### 3. Summary measure analysis
It could be helpful in representing the data to summarize observations into their mean accompanied by standard error.

Summarizing the data

``` {r}
# Time, baseline included
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of RATSL by Group and Time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

```

Plotting the mean weights

``` {r}
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

Figure 3. Mean weights for the three groups in the RATS data

We notice that the three groups have distinctive weights development over time. Their confidence intervals do not overlap. 

Creating a summary data by group and rat with mean as the summary variable (ignoring baseline time 1) and plotting it:

``` {r}

RATS8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATS8S)

# Draw a boxplot of the mean versus treatment
ggplot(RATS8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time excluding baseline")

```



## II - BPRS dataset (Chapter 9 of MABS)