


# Chapter 5: Dimensionality reduction techniques
Mounir Ould Setti
02/12/2019

## 0. Loading the libraries and importing the data
```{r}
library(tidyr)
library(dplyr)
library(stringr)
library(corrplot)
library(GGally)
library(ggplot2)
```

```{r}
human<-read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human2.txt",header=TRUE,sep=",")

```

## 1. Graphical overview of the data, summaries, and distribution
Visualizing the 'human' variables
```{r}
ggpairs(human)
```
Edu2.FM and Edu.Exp have more or less a normal distribution
Labo.FM and Life.Exp are left skewed while GNI, Mat.Mor, Ado.Birth, and Parli.F are right skewed.

Computing the correlation matrix and visualizing it with corrplot
```{r}
cor(human)%>%corrplot
```

There is a strong positive and negative correlation between some variables as the plot shows. For example, Life expectancy has a strong negative correlation with maternal mortality.

## 2. PCA on nonstandardized human data
### Variability of the principal components of nonstandardized human data
```{r}
pca_human <- prcomp(human)
# rounded percetanges of variance captured by each PC
pca_pr <- round(100*(summary(pca_human))$importance[2, ], digits = 1)
# print out the percentages of variance
pca_pr
```
### Biplot of the two principal components of nonstandardized human data
```{r}
# create object pc_lab to be used as axis labels
pc_lab<-paste0(names(pca_pr), " (", pca_pr, "%)")

# draw a biplot
biplot(pca_human, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2])
```


## 3. PCA on standardized human data
```{r}
human <- scale(human)
pca_human <- prcomp(human)
# rounded percetanges of variance captured by each PC
pca_pr <- round(100*(summary(pca_human))$importance[2, ], digits = 1)
# print out the percentages of variance
pca_pr

# create object pc_lab to be used as axis labels
pc_lab<-paste0(names(pca_pr), " (", pca_pr, "%)")

# draw a biplot
biplot(pca_human, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2])
```

The biplots are very different indeed. With unstandardized data, GNI dominates the other variables and the first principal component takes all the variance. The principal components with standardized data are more able to catch the variance in the variables now that the units are equivalent and the distributions are comparable. As a consequence, the first principal component ended up accounting for a reasonable amount of variability (53.6%). The variables Parli.F and Labo.FM take the most from PC2 which accounts for 16% of the data variability while the other variables take more from PC1.

## 4- Interpreting the results

The first principal component seems to be related to factors contributing to the calulation of the Human Development Index
while the second component seems to be related to factors contributing to the calculation of Gender Inequality Index.

Redrawing the plot with personal caption
```{r}
biplot(pca_human, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = "Human Development Index (53.6%)", ylab = "Gender Inequality Index (16.2%)")
```

## 5- The tea dataset

### Exploring and visualizing the dataset

```{r}
library(FactoMineR)
data("tea")
summary(tea)
str(tea)

# visualize the dataset
gather(tea) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free")+ geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

```

### Doing an MCA to certain columns of the data
```{r}
data("tea")
tea_time <- dplyr::select(tea, one_of(c("Tea", "How", "how", "sugar", "where", "lunch")))
# multiple correspondence analysis
mca <- MCA(tea_time,graph = FALSE)
# summary of the model
summary(mca)
# visualize MCA
plot(mca, invisible=c("ind"),habillage = "quali")
```

Dimension 1 accounts for 15% of the variability in the data and Dimension 2 accounts for 14%.